Liveness-Based Garbage Collection

### 2014-07-13 paper/sections/abstract.tex
organization added

%------------------------------------------------------------%
\documentclass{llncs}
\pagestyle{plain}
\usepackage{longtable} 
%------------------------------------------------------------%
%- common stuff starts here....
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
%\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
%\usepackage{color}
%\usepackage{xcolor}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
%\usepackage{pst-all}
\usepackage{bcprules}
%\usepackage{subfigure}
%\usepackage{algorithmic}
%\usepackage{anysize}     
\input{fh_defs}
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}

\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

\begin{document}
%  \title{Heap  Reference  Analysis  for Functional  Programs}
\title{Liveness-Based Garbage Collection}

% make a proper TOC despite llncs
\setcounter{tocdepth}{2}
\makeatletter
\renewcommand*\l@author[2]{}
\renewcommand*\l@title[2]{}
\makeatletter

  \author{Rahul Asati\inst{1} \and Amitabha Sanyal\inst{1} \and Amey Karkare\inst{2}
    \and   Alan Mycroft\inst{3}}
  \institute{IIT Bombay, Mumbai 400076, India\\
    \email{$\{$rahulasati,as$\}$@cse.iitb.ac.in}, \and 
             IIT Kanpur, Kanpur 208016, India\\
    \email{karkare@cse.iitk.ac.in}, \and 
             Computer Laboratory, University of Cambridge, CB3 0FD, UK\\
    \email{alan.mycroft@cl.cam.ac.uk}}
\maketitle

\begin{abstract}
Current garbage collectors  leave much heap-allocated data uncollected
because they preserve data {\em reachable} from a root set.
% However, it is only necessary to preserve {\em live} data---a subset of
% reachable data.
However, only  {\em live} data---a subset of  reachable data---need be
preserved.

@@

Experiments  confirm the  expected performance  benefits---increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.
\end{abstract}

%------------------------------------------------------------%

%% \begin{figure}
%% \tableofcontents       %%%%%%% BEWARE: find the \setcounter{page}{2}
%%                        %%%%%%% and comment it out when deleting this TOC.
%% \vspace{5cm}

%% \section*{Page Budget}
%% CC'2014 and ESOP'2014 have a limit of 20 pages.
%% A draft page budget is:
%% \begin{itemize}
%% \item 1. abstract/introduction (3 pages)
%% {\color{red}Currently 2.5 pages.}
%% \item 2. core language syntax and semantics (2 pages).
%% {\color{red}Currently 2 pages.}
%% \item 3. liveness analysis and correctness (3 pages)
%% {\color{red}Currently 3.8 pages.}
%% \item 4. practical liveness computation (6 pages)
%% {\color{red}Currently 6.2 pages.}
%% \\ I think this is where should do all 3 of: the context-insensitive analysis
%% (one which has a summary demand for each function), then the
%% solving like differential equations, then using Mohri-Nederhof to
%% get a NFA for each (call/cons) program point.
%% \item 5. experimental framework and results (4 pages)
%% \item 6. (optional) monotonicity of GC (0.5 page)---our technique can never
%% cause more collections.
%% \item 7. conclusions/acknowledgements/bibliography (1.5 page)
%% \end{itemize}
%% \end{figure}

\section{Introduction}
\label{sec:intro}

Most modern  programming languages support dynamic  allocation of heap
data.  Static  analysis of heap data  is much harder  than analysis of
static   and   stack   data.    Garbage   collectors,   for   example,
conservatively  approximate  the liveness  of  heap  objects by  their
reachability  from a  set of  memory  locations called  the {\em  root
  set\/}.  Consequently, many objects  that are reachable but not live
remain  uncollected, causing  a  larger-than-necessary memory  demand.
This is confirmed  by empirical studies on Haskell~\cite{rojemo96lag},
Scheme~\cite{karkare06effectiveness}                                and
Java~\cite{shaham02estimating} programs.

Here we consider a first-order  pure functional language and propose a
liveness  analysis  which  annotates  various program  points  with  a
description  of variables and  fields whose  object references  may be
dereferenced  in the future.   The garbage  collector then  only marks
objects pointed by live references and leaves other, merely reachable,
objects  to  be  reclaimed.    (Although  not  strictly  necessary,  a
collector  would  normally {\em  nullify}  dead  variables and  fields
rather than leaving dangling  references.)  Since there are fewer live
objects   than   reachable   objects,   more  memory   is   reclaimed.
Additionally, since  the collector traverses a smaller  portion of the
heap, the time spent for each collection is also smaller.  The work is
presented in  the context of a  stop-the-world non-incremental garbage
collector (mark-and-sweep,  compacting or  copying) for which  we also
show a  monotonicity result: that  our technique can never  cause more
garbage  collections  to  occur   in  spite  of  changing  the  rather
unpredictable  execution  points   at  which  collections  occur.   We
anticipate that our technique  is applicable to more modern collectors
(generational,  concurrent, parallel),  but leave  such  extensions to
future work.

We first define a {\em fully context-sensitive} (in the sense that its
results  are unaffected  by function  inlining) liveness  analysis and
prove it  correct.  However, fully context-sensitive  methods often do
not scale,  and this analysis would  also require us  to determine, at
run-time,  the internal  liveness of  a  function body  at each  call.
Hence,   similarly   to   the   0-CFA   approach,   we   determine   a
context-independent {\em summary} of  liveness for each function which
safely   approximates   the   context-dependence   of   all   possible
calls~\cite{chatterjee99relevant,cherem07fast,lee05static}.       (Note
that  an intraprocedural context-insensitive  method which  assumes no
information  about function  callers would  be too  imprecise  for our
needs.)   In essence  our approach  sets up  interprocedural data-flow
equations for the liveness summaries  of functions and shows how these
can be solved symbolically  as context-free grammars (CFGs)\@.  We can
then determine  a CFG  for each program  point; these are  then safely
approximated with  finite-state automata  which are encoded  as tables
for each program point.  For  garbage collection purposes only {\em GC
  points}  (program points  representing  calls, to  $\CONS$  or to  a
user-function) need to be stored.
% which may call eventually call $\CONS$, need to be stored.

We  previously proposed  an intraprocedural  method for  heap liveness
analysis   for   a   Java-like   language~\cite{khedker07heap}   which
statically inserted  statements nullifying dead  references to improve
garbage collection; by  contrast nullification here occurs dynamically
(which can work better with  aliasing) when the garbage collector acts
on  liveness  annotations  to  avoid traversing  dead  references.   A
workshop    paper~\cite{karkare07liveness}    outlined    the    basic
0-CFA-style-summary  interprocedural  approach  to  functional-program
liveness  analysis.   The  current  paper adds  the  context-sensitive
analysis and better formalization along with experimental results.
%   to  substantiate our ideas.

\kern -3pt   %% AM: a very small tweak to get back to 2 pages for this
             %% section.
%% AM: rewrite the next paragraph to save 3-4 words and remove this \kern.

%------------------------------------------------------------%
\setcounter{page}{2}
\subsubsection{Motivating Example}
\label{sec:motiv}

Figure~\ref{fig:mot-example}(a) shows  an example program.   The label
$\pi$ of an expression $e$  denotes a program point.  During execution
of  the program, it  represents the  instant of  time just  before the
evaluation  of  $e$.  
%%While we discuss liveness, it  will also be important to consider an
%%instance of time  just after the execution of  an expression, though
%%we shall not be required to mark it with a label.
% memory as  a (possibly  disconnected) directed
% acyclic   graph.\footnote{Since  the   language   under  consideration
%   (Section~\ref{sec:defs}) does not  have any imperative features, the
%   heap cannot  have cycles.}
% Here all data  is assumed to  be boxed, i.e.\
% stored in heap and accessed through references.
We  view the  heap  as a graph.  
Nodes in the heap, also  called  ($\CONS$) {\em  cells\/} contain $\CAR$ and $\CDR$ {\em fields}
containing values.  Edges in the graph are {\em references} and
emanate from {\em variables} or fields.
Variable and field values may also be atomic values ($\NIL$, integers etc.)
While it is convenient to box these in diagrams, our presented
analysis treats them as non-heap values.

Figure~\ref{fig:mot-example}(b)  shows the heap at $\pi$.
The edges shown  by thick arrows are those  which are made live by the
program. In addition,  assuming that
the value of any reachable part of the program result may be explored
or printed, the edges marked by thick dashed arrows are also live.
A cell  is  marked and preserved  during  garbage collection,  only  if it  is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if  a garbage collection takes
% place   at    the   time   represented   by   the    heap   shown   in
at $\pi$ with the    heap   shown   in
Figure~\ref{fig:mot-example}(b), only the cells $\pw$ and $(\CDR~ \pw)$, along with $(\CAR~(\CDR~\pw))$ and all cells reachable from it,
will be marked and preserved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

\begin{figure}[t!]
%%%%{\color{Myblue}
\begin{picture}(100,130)(0,-90)
%  \begin{center}
    \begin{tabular}{cc}
\begin{boxedminipage}{.5\textwidth}
      {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UFL\ \hspace*{-.31\TAL} (\DEFINE\ (\append\  \lista\ \listb)
	  \UNL{0}  (\SIF~(\NULLQ \ \lista)
	       \listb
	  \UNL{1}      (\CONS\ (\CAR\  \lista)
%          \UNL{2}          (\CAR\  \lista)
	  \UNL{2}\;\;\;\;\;          (\append\ (\CDR\  \lista)
          \listb))))
          \UNL{0}
	  \UNL{0} \hspace*{-.49\TAL} (\LET\ z  $\leftarrow$(\CONS\ (\CONS\ $4$ (\CONS\ $5$ \NIL))
	  \UNL{3}\ \ \ \  (\CONS\ $6$ \NIL)) \IN
	  \UNL{0} (\LET\ y  $\leftarrow$ (\CONS\ $3$ \NIL) \IN
	  \UNL{1}   (\LET\ w $\leftarrow$ (\append\ y z)\ \IN
	  \UNL{1}               $\,\,\,\,\,\,\,\,\pi$:(\CAR\ (\CDR\ w)))))
	\end{uprogram}
      }} 
\end{boxedminipage}
      &
      \raisebox{-25mm}{\scalebox{.75}{
	%%%%%%%%%%%%%%%%%%%%%Uday's stuff%%%%%%%%%%%%%%%%%%%%%%%%%
      \psset{unit=1mm}
      \psset{linewidth=.3mm}
      \begin{pspicture}(0,-5)(70,60)
	%\psframe(0,0)(73,60)
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{o}{origin}{13}{50}{\TwoCells{o1}{o2}}
	\putnode{a}{o}{-10}{-15}{\psframebox{3}}
%	\putnode{b}{o}{10}{-15}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o1}{a}
	%\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	%      fillcolor=white]{$\times$}}}
        \putnode{b}{o}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{o2}{b}
	%\ncline[offsetB=-.5,nodesepB=.1]{->}{o2}{b}
	\putnode{y}{o}{-14}{8}{\psframebox[linestyle=none,framesep=.5]{y}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120]{->}{y}{o}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
		  fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{c}{o}{25}{0}{\TwoCells{c1}{c2}}
	\putnode{d}{c}{10}{-10}{\TwoCells{d1}{d2}}
	\putnode{e}{d}{-13}{-12}{\TwoCells{e1}{e2}}
	\putnode{f}{d}{13}{-12}{\TwoCells{f1}{f2}}
	\ncline[nodesepB=-.5]{*->}{c2}{d}
	\ncline[nodesepB=-.5,linewidth=.7]{->}{c2}{d}
	\nccurve[ncurv=1,angleA=270,angleB=330]{*->}{c1}{a}
	\aput[-3.5](.2){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70]{*->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=240,angleB=70,linewidth=.7]{->}{d1}{e}
	\nccurve[nodesepB=-.5,angleA=300,angleB=110]{*->}{d2}{f}
	\aput[-3.5](.5){\scalebox{1.2}{\psframebox[framesep=.2,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	\putnode{w}{c}{-8}{8}{\psframebox[linestyle=none,framesep=.2]{w}}
	\putnode{ww}{c}{15}{8}{\psframebox[linestyle=none,framesep=.2]{z}}
	\nccurve[nodesepB=-.2,angleA=330,angleB=120,linewidth=.7]{->}{w}{c}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{g}{e}{-8}{-12}{\psframebox{4}}
	\putnode{h}{e}{8}{-14}{\TwoCells{h1}{h2}}
	\putnode{i}{f}{-8}{-11}{\psframebox{6}}
%	\putnode{j}{f}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{e1}{e1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e1}{g}
        %here
        \putnode{j1}{f}{0}{-3}{\psframebox[linestyle=none,framesep=0]{\scalebox{.63}{\nilfigure}}}
        \putnode{j2}{h}{0}{-3}{\psframebox[linestyle=none,framesep=.5]{\scalebox{.63}{\nilfigure}}}

%        \aput[-3.2](.6){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here
	\ncline[offsetB=-.5,nodesepB=-.3]{*-}{e2}{e2}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{e2}{h} 
%	\aput[-3.2](.5){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
%	      fillcolor=white]{$\times$}}} %and here

	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f1}{i}
	\ncline[offsetB=-.5,nodesepB=.1]{*->}{f2}{j}
	\nccurve[nodesepB=-.2,angleA=270,angleB=90]{->}{ww}{d}
	\aput[-3.2](.4){\scalebox{1.2}{\psframebox[framesep=.1,linestyle=none,fillstyle=solid,
	      fillcolor=white]{$\times$}}}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\putnode{k}{h}{-8}{-11}{\psframebox{5}}
%	\putnode{l}{h}{8}{-11}{\psframebox[linestyle=none,framesep=.5]{\NIL}}
	\ncline[offsetB=-.5,nodesepB=.1]{*-}{h1}{h1}
	\ncline[linestyle=dashed,offsetB=-.5,nodesepB=.1,linewidth=.7]{->}{h1}{k}
%	\ncline[offsetB=-.5,nodesepB=.1]{*->}{h2}{l}
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \end{pspicture}}} \\
      (a) Example program.&
      \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
	(b) Memory graph at $\pi$. Thick edges denote \\
        {\white (b) }live links. Traversal stops at edges marked\\
        {\white (b) }$\times$  during garbage collection.
      \end{tabular}}
    \end{tabular}
\end{picture}

\kern -3ex

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure}

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational}   describes    the
% operational semantics  of the  language.
Liveness analysis is described in
Section~\ref{sec:liveness} followed by a sketch of a correctness
proof relative to a non-standard semantics.
Section~\ref{sec:computing} shows how to encode liveness as
finite-state automata.
Section~\ref{sec:experiments}   reports   experimental   results and
Section~\ref{sec:lgc-always-better} proves that a liveness based
collector can never do more garbage collections than a reachability
based collector. 
% and Section~\ref{sec:conclusion} concludes the paper.

\section{The target language---syntax and semantics}
\label{sec:defs}
We let $x$, $y$, $z$ range over variables, $f$ over user-functions and $p$ over primitive
functions ($\CONS$, $\PRIM$ etc.).
The syntax  of our language is  shown in Figure~\ref{fig:lang-syntax};
it has eager semantics and restricts programs to be in 
Administrative  Normal Form (ANF)~\cite{chakravarty03perspective}
where all actual parameters to functions are variables.
This restriction does not affect expressibility (and indeed we feel free
to ignore it in examples when inessential), but simplifies the analysis formulation.
Additionally, as in the three-address instruction form familiar from compiler texts,
it forces each temporary to be named and function calls to be serialized
(necessary to get an unambiguous definition of liveness).
We further require that each variable in a program is distinct, so that no scope shadowing
occurs---this simplifies proofs of soundness.
In this formulation expressions: either perform a test ($\SIF$), make a computation step ($\LET$)
or return ($\RETURN$).
The $\RETURN$ keyword is logically redundant, but we find it clarifies the semantics and analysis.

The body of the program is
the  expression denoted  by  $e_\mainpgm$; for analysis purposes
it is convenient
to regard $e_\mainpgm$ as part of a function definition
$(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
% As  mentioned earlier,
We write $\pi\!:\!e$ to associate the  label $\pi$ (not part of the language syntax) with the program point just  before expression $e$.

% \paragraph{Inlining.}
In spite of the ANF restrictions it is still possible
to inline non-recursive functions (a fact we use to prove the safety
of liveness analysis).  A user-function call
$(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)$
to a function defined
(after renaming its formals and locals to be disjoint from existing variables)
by $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$
is replaced by a sequence of $\LET$'s of the form
$z_i\leftarrow (\ID\,\,y_i)$ followed by
the body $e_f$ but with its $(\RETURN\,\,w)$ expressions replaced by
$(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.

\begin{figure}[t!]
\footnotesize
\begin{eqnarray*}
   p \in \mathit{Prog} & ::= & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{7em} \mbox{\em --- program}\\
    d \in Fdef & ::= & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots \,\,x_n)\,\,
    e) 
    \hspace{1.7em} \mbox{\em --- function definition} \\
e \in \mathit{Expr} & ::= &
\left\{\begin{array}{ll@{\hspace{4em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \mbox{\em --- conditional} \\ 
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) && \mbox{\em --- let binding} \\
       (\RETURN\,\, x) && \mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{Stmt} & ::= &
\left\{\begin{array}{lr@{\hspace{1em}}l}
       k && \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) && \mbox{\em --- constructor} \\ 
       (\CAR\,\, x) &  (\CDR\,\, x) & \mbox{\em --- selectors} \\ 
       (\NULLQ\,\, x) & (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester and generic arithmetic} \\ 
%        (\NULLQ\,\, x) && \mbox{\em ---  tester} \\ 
       (\ID\,\, x) && \mbox{\em ---  identity function (for inlining)} \\ 
%        (\PRIM\,\, x_1\,\, x_2) && \mbox{\em --- generic arithmetic} \\ 
       \multicolumn{2}{l}{(f\,\, x_1\,\,\ldots\,\, x_n)} 
            & \mbox{\em --- function application} 
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\subsubsection{Semantics}
We now give an operational
semantics for  our language.  Later, a refinement of
the  operational semantics,  which we  call minefield  semantics, will
serve to prove liveness analysis correct.
We give a small-step semantics because, unlike big-step semantics,
correctness for non-terminating programs does not need special treatment.
We start with the domains used by the semantics:
\[
\begin{array}{rlcl@{\hspace{2em}}l}
% \begin{eqnarray*}
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Loc}& \mbox{-- Values}\\
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow {Val} &
\mbox{-- Environment} \\ 
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow ({Val\times
  Val}) & \mbox{-- Heap}
% \end{eqnarray*}
\end{array}
\]
Here $\mathit{Loc}$ is countable set of locations which hold $\CONS$ cells.
A value is either a number, the empty list $\NIL$, or a location  $\ell$.
Our liveness analysis does not track numeric values, and thus is neutral
as to whether these are boxed or represented as immediates.
An environment is a finite mapping from variables to values, and a heap a finite mapping
from locations to pairs of values.
% Given  a program,  an  environment is  a
% mapping from  the set  of variables of  the program  $\mathit{Var}$ to
% values in $\mathit{Val}$.    The heap is a mapping
% from  locations  to pairs of  values.
Finally, $\stk$ is a stack
(using $\bullet$ for push and $[\,]$ for empty stack)
of frames of unfinished function calls.  A
frame is a triple $(e,x,\rho)$ representing the call site
$
(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n) \,\, \IN\,\, e)
$
being evaluated in environment $\rho$.
Frames can also be viewed as continuations, in this view
the ({\sc ord-return}) 
 rule in the small-step operational semantics (Figure~\ref{fig:lang-opsem-small}) invokes them.

The semantics 
of statements $s$ are given by the judgement form
$\rho, \heap, s \rightsquigarrow \heap', v$
and those for expressions $e$ by the form
$\rho, \stk, \heap, e \rightarrow \rho', \stk', \heap', e'$.
The start state is $(\{\},[\,], \{\}, e_\mainpgm)$ and the
program terminates successfully with result value $\rho(x)$
on reaching the halt state $(\rho,[\,], \heap, (\RETURN\ x))$

Notation:  we write $\rho[x \mapsto v]$ for the environment which is
as $\rho$ but has value $v$ at $x$.
We also write $[\myvec{x} \mapsto \myvec{v}]$ which respectively
has values $v_1, \ldots, v_n$ at $x_1, \ldots, x_n$ and write
$[\myvec{x} \mapsto \rho(\myvec{y})]$ when 
$v_1, \ldots, v_n$ are $\rho(y_1), \ldots, \rho(y_n)$.

\paragraph{Stuck states.} Note that certain forms of $e$ do not reduce
with $\rightarrow$ (perhaps because $\rightsquigarrow$ could not reduce
a contained $s$).  Some of these we eliminate syntactically, e.g.\
ensuring all variables and functions are defined and are called with
the correct number of parameters.
Others include $(\CDR\ \NIL), (\CAR\ 3), (+\ \NIL\ 4)$ and $(\SIF\ \NIL\ e_1\ e_2)$.
All but the first can be eliminated with a static type system but,
treating our program as dynamically typed, we regard all these
as stuck states.

% A program terminates successfully if it reaches the halt state
% in Figure~\ref{fig:lang-opsem-small}.
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

\begin{figure}[t!]  
% (Random text to ensure proper line-spacing (problem with infrule macro?).)

\footnotesize
\begin{minipage}{0.335\textwidth}
\infrule[ord-const]
        {}
        {\rho, \heap, \pp{\pi}{k}
          \rightsquigarrow \heap, k}
\end{minipage}\begin{minipage}{0.655\textwidth}
\infrule[ord-cons]
      {
\mbox{$\ell \not\in \mbox{dom}(\heap)$ is a fresh location}
}
      {\rho, \heap, \pp{\pi}{(\CONS\,\, x\ y)}
        \rightsquigarrow \heap[\ell\mapsto(\rho(x),\rho(y))],
        \ell}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-car]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CAR\,\, x)}
          \rightsquigarrow \heap, v_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-cdr]
        {\heap(\rho(x)) = (v_1, v_2)}
        {\rho, \heap, \pp{\pi}{(\CDR\,\, x)}
          \rightsquigarrow \heap, v_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
\infrule[ord-id]
      {
}
      {\rho, \heap, \pp{\pi}{(\ID\,\, x)}
        \rightsquigarrow \heap, \rho(x)}
\end{minipage}\begin{minipage}{0.595\textwidth}
\infrule[ord-prim]
      {\rho(x) \in \mathbb{N} \andalso \rho(y)\in \mathbb{N}}
      {\rho, \heap, \pp{\pi}{(\PRIM\,\, x\ y)}
        \rightsquigarrow \heap, \rho(x) + \rho(y)}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-null-false]
\infrule[]
        {\rho(x)\not= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 0}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-null-true]
\infrule[ord-null]
        {\rho(x)= \NIL
        }{\rho, \heap, \pp{\pi}{(\NULLQ\,\, x)}
          \rightsquigarrow \heap, 1}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.395\textwidth}
% \infrule[ord-if-true]{
\infrule[]{
  \rho(x) \in \mathbb{N}\setminus\{0\}
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_1}
\end{minipage}\begin{minipage}{0.595\textwidth}
% \infrule[ord-if-false]{
\infrule[ord-if]{
  \rho(x) = 0
}{
  \rho,\stk, \heap,
  \pp{\pi}{(\SIF\ x\ \pp{\pi_2}{e_1}\ \pp{\pi_3}{e_2})}
  \longrightarrow \rho,\stk, \heap, e_2}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-nonfn]{
%  \rho,\stk, \heap, \pp{\pi_1}{s} \longrightarrow \rho,\stk, \heap', v_1 
  \rho, \heap, \pp{\pi_1}{s} \rightsquigarrow \heap', v 
  \andalso
  \mbox{$s$ is not
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
  \rho[x \mapsto v], \stk, \heap', \pp{\pi_2}{e}}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-let-fncall]{
  \mbox{$s$ is 
$\pp{\pi}{(f\,\, \pp{\pi_1}{y_1} \ldots \pp{\pi_n}{y_n})}$
       }
\andalso
\mbox{$f$ defined as $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \pp{\pi_f}{e_f})$}
}{
  \rho,\stk, \heap,  \pp{\pi}{(\LET\,\, x
    \leftarrow  \pp{\pi_1}{s}\,\, \IN\,\,
    \pp{\pi_2}{e})}
  \longrightarrow
% [z_1\mapsto \rho(y_1), \ldots, z_n\mapsto \rho(y_n)], 
[\myvec{z} \mapsto \rho(\myvec{y})],
  \,(\rho,x,e)\bullet\stk,\, \heap, e_f
}
\end{minipage} \\[0.5ex]

\begin{minipage}{0.990\textwidth}
\infrule[ord-return]{
}{
  \rho, \,(\rho',x',e')\bullet\stk,\, \heap,
  \pp{\pi}{(\RETURN\ x)}
  \longrightarrow \rho'[x' \mapsto \rho(x)], \stk, \heap, e'}
\end{minipage}

% \normalsize
% Note that $\rho,[\,], \heap, \pp{\pi}{(\RETURN\ x)}$
% is the halt state (with result value $\rho(x)$).

  \caption{The small-step operational semantics}
  \label{fig:lang-opsem-small}
\figrule
\normalsize
\end{figure}

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point

\section{Liveness}\label{sec:liveness}

In classical liveness analysis a variable is either `live' (its value
may be used in future computation) or `dead' (definitely not used).
Semantically, a variable is dead at a given program point
if arbitrary changes to its value have no effect on the computation.
Later we will use $\bot$ to represent a value which `explodes' when
it is used in a computation; dead variables can safely have their
value replaced with $\bot$.
For heap-allocated data  we need a richer model of liveness
in that both variables and fields of $\CONS$ cells may be dead or live.
Using $\acar$, $\acdr$ to represent access using $\CAR$, $\CDR$ respectively,
liveness of the structure reachable from a variable is a set of
{\em access paths}
which we represent as a subset of $\{\acar,\acdr\}^\ast$, and use conventional
grammar notation.
Thus the liveness of $x$ being
$\{\acdr\acar, \acdr\acdr\acar\}$ means that future computation can
only refer to the second and third members of $x$ considered as a list.
Semantically, access paths are prefix-closed, as accessing a field
requires accessing all the paths from the variable to the field,
and hence the above liveness is properly written 
$\{\epsilon, \acdr, \acdr\acar, \acdr\acdr, \acdr\acdr\acar\}$.
The classical notions of a scalar variable being live or dead correspond to
$\{\epsilon\}$ and $\{\}$.

The overall liveness (also written {\em liveness environment} for emphasis)
at a program point is conceptually a mapping from
variables to subsets of $\{\acar,\acdr\}^\ast$, but we often abuse
notation, for example writing $\{x.\acar\acdr, x.\acdr, y.\epsilon\}$
instead of the map
$[x \mapsto\{\epsilon,\acar,\acar\acdr,\acdr\}, y\mapsto\{\epsilon\}, z\mapsto\{\}]$.
Analogously to classical liveness, the liveness at program point $\pi$
in $\pi:e$ is the liveness just before executing $e$.

>>>>

Current garbage collectors  leave much heap-allocated data uncollected
because they preserve data {\em reachable} from a root set.
% However, it is only necessary to preserve {\em live} data---a subset of
% reachable data.
However, only  {\em live} data---a subset of  reachable data---need be
preserved.

@@

Experiments  confirm the  expected performance  benefits---increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.
